Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-02-01 paper/main.tex
edits through discussion

Although a single lecture may be organized around a single broad theme at a
coarse scale, at a finer scale each moment of a lecture typically covers a
narrower range of content. We wondered whether a text embedding model trained
on the lectures' transcripts might capture some of this finer scale content.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether the text embeddings could be used to
automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors from each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also qualitatively examined the
best-matching intervals for each question by comparing the text
of the question to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question probed and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

@@

We developed a computational framework that uses short multiple choice quizzes
to provide nuanced insights into what learners know and how their knowledge
changes with training. First, we show that our approach can automatically match
up the conceptual content of individual questions with the corresponding
moments in lecture videos when those concepts were presented
(Fig.~\ref{fig:question-correlations}). Next, we demonstrate how we can
estimate moment-by-moment ``knowledge traces'' that reflect how much knowledge
participants have about each video's content before and after watching each
lecture video (Fig.~\ref{fig:knowledge-timeseries}). We also show that these
knowledge estimates can generalize to held-out questions
(Fig.~\ref{fig:predictions}). Finally, we use our framework to construct visual
maps that provide snapshot estimates of how much participants know about any
concept within the scope of our text embedding model, and how much their
knowledge changes with training (Fig.~\ref{fig:knowledge-maps}).

Over the past several years, the global pandemic has forced many educators to
teach remotely~\citep{MoseEtal21, ShimLee20, KawaEtal21, Whal20}. This change
in world circumstances is happening alongside (and perhaps accelerating)
geometric growth in the availability of high quality online courses on
platforms such as Khan Academy~\citep{Khan04}, Coursera~\citep{Youn12},
EdX~\citep{Kolo13}, and others~\citep{RhoaEtal13}. Continued expansion of the
global internet backbone and improvements in computing hardware have also
facilitated improvements in video streaming, enabling videos to be easily
downloaded and shared by large segments of the world's population. This
exciting time for online course instruction provides an opportunity to
re-evaluate how we, as a global community, educate ourselves and each other.
For example, we can ask: what makes an effective course or training program?
Which aspects of teaching might be optimized or automated? How and why do
learning needs and goals vary across people? How might we lower barriers to
achieving a high quality education?

Alongside these questions, there is a growing desire to extend existing
theories beyond the domain of lab testing rooms and into real
classrooms~\citep{Kauf03}. In part, this has led to a recent resurgence of
``naturalistic'' or ``observational'' experimental paradigms that attempt to
better reflect more ethologically valid phenomena that are more directly
relevant to real-world situations and behaviors~\citep{NastEtal20}. In turn,
this has brought new challenges in data analysis and interpretation. A key step
towards solving these challenges will be to build explicit models of real-world
scenarios and how people behave in them (e.g., models of how people learn
conceptual content from real-world courses, as in our current study). A second
key step will be to understand which sorts of signals derived from behaviors
and/or other measurements~\citep[e.g., neurophysiological data; ][]{NguyEtal22,
MeshEtal20, PoulEtal17, BeviEtal19, DikkEtal17} might help to inform these
models. A third major step will be to develop and employ reliable ways of
evaluating the complex models and data that are a hallmark of naturalistic
paradigms.

Beyond specifically predicting what people \textit{know}, the fundamental ideas
we develop here also relate to the notion of ``theory of mind'' of other
individuals~\citep{GoldWinn12, KansEtal15, Melt11}. Considering others' unique
perspectives, prior experiences, knowledge, goals, etc., can help us to more
effectively interact and communicate~\citep{ShaoEtal18, StepBaer06, Ratk18}.
One could imagine future extensions of our work (e.g., analogous to the
knowledge and learning maps shown in Fig.~\ref{fig:knowledge-maps}), that
attempt to characterize how well-aligned different people's knowledge bases or
backgrounds are. In turn, this might be used to model how knowledge (or other
forms of communicable information) flows not just between teachers and
students, but between friends having a conversation, individuals out on a first
date, participants at a business meeting, doctors and patients, experts and
non-experts, political allies or adversaries, and more. For example, the extent
to which two people's knowledge maps ``match'' or ``align'' in a given region
of text embedding space might serve as a predictor of how effectively the
people will communicate about the corresponding conceptual content.

Ultimately, our work suggests a rich new line of questions about the geometric
``form'' of knowledge, how knowledge changes over time, and how we might map
out the full space of what an individual knows. Our finding that detailed
estimates about knowledge may be obtained from short quizzes shows one way that
traditional approaches to evaluation in education might be extended. We hope
that these advances might help pave the way for new ways of teaching or
delivering educational content that are tailored to individual students'
learning needs and goals.

>>>>

Although a single lecture may be organized around a single broad theme at a
coarse scale, at a finer scale each moment of a lecture typically covers a
narrower range of content. We wondered whether a text embedding model trained
on the lectures' transcripts might capture some of this finer scale content.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether the text embeddings could be used to
automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors from each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also qualitatively examined the
best-matching intervals for each question by comparing the text
of the question to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question probed and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

@@

We developed a computational framework that uses short multiple-choice quizzes
to gain nuanced insights into what learners know and how their knowledge
changes with training. First, we show that our approach can automatically match
the conceptual knowledge probed by individual quiz questions to the corresponding
moments in lecture videos when those concepts were presented
(Fig.~\ref{fig:question-correlations}). Next, we demonstrate how we can
estimate moment-by-moment ``knowledge traces'' that reflect the degree of knowledge
participants have about each video's time-varying content, and capture temporally 
specific increases in knowledge after viewing each lecture (Fig.~\ref{fig:knowledge-timeseries}).
 We also show that these knowledge estimates can generalize to held-out questions
(Fig.~\ref{fig:predictions}). Finally, we use our framework to construct visual
maps that provide snapshot estimates of how much participants know about any
concept within the scope of our text embedding model, and how much their
knowledge changes with training (Fig.~\ref{fig:knowledge-maps}).

Over the past several years, the global pandemic has forced numerous educators to
teach remotely~\citep{MoseEtal21, ShimLee20, KawaEtal21, Whal20}. This change
in world circumstances is happening alongside (and perhaps accelerating)
geometric growth in the availability of high quality online courses on
platforms such as Khan Academy~\citep{Khan04}, Coursera~\citep{Youn12},
EdX~\citep{Kolo13}, and others~\citep{RhoaEtal13}. Continued expansion of the
global internet backbone and improvements in computing hardware have also
facilitated improvements in video streaming, enabling videos to be easily
shared and viewed by large segments of the world's population. This
exciting time for online course instruction provides an opportunity to
re-evaluate how we, as a global community, educate ourselves and each other.
For example, we can ask: what makes an effective course or training program?
Which aspects of teaching might be optimized and/or augmented by automated tools? How and why do
learning needs and goals vary across people? How might we lower barriers to
achieving a high-quality education?

Alongside these questions, there is a growing desire to extend existing
theories beyond the domain of lab testing rooms and into real
classrooms~\citep{Kauf03}. In part, this has led to a recent resurgence of
``naturalistic'' or ``observational'' experimental paradigms that attempt to
better reflect more ethologically valid phenomena that are more directly
relevant to real-world situations and behaviors~\citep{NastEtal20}. In turn,
this has brought new challenges in data analysis and interpretation. A key step
towards solving these challenges will be to build explicit models of real-world
scenarios and how people behave in them (e.g., models of how people learn
conceptual content from real-world courses, as in our current study). A second
key step will be to understand which sorts of signals derived from behaviors
and/or other measurements (e.g., neurophysiological data;~\citealp{NguyEtal22,
MeshEtal20, PoulEtal17, BeviEtal19, DikkEtal17}) might help to inform these
models. A third major step will be to develop and employ reliable ways of
evaluating the complex models and data that are a hallmark of naturalistic
paradigms.

Beyond specifically predicting what people \textit{know}, the fundamental ideas
we develop here also relate to the notion of ``theory of mind'' of other
individuals~\citep{GoldWinn12, KansEtal15, Melt11}. Considering others' unique
perspectives, prior experiences, knowledge, goals, etc., can help us to more
effectively interact and communicate~\citep{ShaoEtal18, StepBaer06, Ratk18}.
One could imagine future extensions of our work (e.g., analogous to the
knowledge and learning maps shown in Fig.~\ref{fig:knowledge-maps}), that
attempt to characterize how well-aligned different people's knowledge bases or
backgrounds are. In turn, this might be used to model how knowledge (or other
forms of communicable information) flows not just between teachers and
students, but between friends having a conversation, individuals on a first
date, participants at a business meeting, doctors and patients, experts and
non-experts, political allies or adversaries, and more. For example, the extent
to which two people's knowledge maps ``match'' or ``align'' in a given region
of text embedding space might serve as a predictor of how effectively they 
will be able to communicate about the corresponding conceptual content.

Ultimately, our work suggests a rich new line of questions about the geometric
``form'' of knowledge, how knowledge changes over time, and how we might map
out the full space of what an individual knows. Our finding that detailed
estimates about knowledge may be obtained from short quizzes shows one way that
traditional approaches to evaluation in education may be extended. We hope
that these advances might help pave the way for new approaches to teaching or
delivering educational content that are tailored to individual students'
learning needs and goals.
